---
title: "Data Analysis Assignment #2 (75 points total)"
author: "Sharma, Sachin"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```

##Data Analysis #2

```{r analysis_setup1, message = FALSE, warning = FALSE}

# Perform the following steps to start the assignment.
 
# 1) Load/attach the following packages via library():  flux, ggplot2, gridExtra, moments, rockchalk, car.
# NOTE:  packages must be installed via install.packages() before they can be loaded.

library(dplyr)
library(flux)
library(ggplot2)
library(gridExtra)
library(knitr)
library(rockchalk)
library(tidyverse)
library(RColorBrewer)
library(magrittr)
my_palette <- brewer.pal(name="GnBu", 9)[4:9]

# 2) Use the "mydata.csv" file from Assignment #1 or use the file posted on the course site.  Reading
# the files into R will require sep = "" or sep = " " to format data properly.  Use str() to check file
# structure.

mydata <- read.csv("mydata.csv", sep = ",", stringsAsFactors = TRUE)
# mydata <- read.csv(file.path("c:...", "mydata.csv"), sep = ",")
# mydata <- read.csv(file.path("c:/Rabalone/", "mydata.csv"), sep = ",")

str(mydata)

```

### Test Items starts from here - There are 10 sections - total of 75 points ##############

***#### Section 1: (5 points) ####***

(1)(a) Form a histogram and QQ plot using RATIO. Calculate skewness and kurtosis using 'rockchalk.' Be aware that with 'rockchalk', the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_1a, fig.width = 12}

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO  <- mydata$SHUCK / mydata$VOLUME
s  <- rockchalk::skewness(mydata$RATIO)
k  <- rockchalk::kurtosis(mydata$RATIO, excess = FALSE)
k2 <- rockchalk::kurtosis(mydata$RATIO)

par(mfrow = c(1, 2), oma = c(0, 0, 2, 0))
hist(mydata$RATIO, col = "dodgerblue", xlab = "Ratio",
     main = "Histogram of Ratio (Shuck/Vol)")
qqnorm(mydata$RATIO, main = "QQ Plot",
       col = ifelse(mydata$RATIO %in% c(boxplot.stats(mydata$RATIO)$out), 
                    "orange", "green"))
qqline(mydata$RATIO)
txt <- c("Skewness:", round(s, 2), 
         "Kurtosis:", round(k, 2), 
         "Excess Kurtosis:", round(k2, 2))
mtext(paste(txt, sep=" ", collapse="  "), outer = TRUE, cex = 1)

```

(1)(b) Tranform RATIO using *log10()* to create L_RATIO (Kabacoff Section 8.5.2, p. 199-200). Form a histogram and QQ plot using L_RATIO. Calculate the skewness and kurtosis. Create a boxplot of L_RATIO differentiated by CLASS.

```{r Part_1b, fig.width = 12, fig.height = 8}

mydata$L_RATIO <- log10(mydata$RATIO)

s  <- rockchalk::skewness(mydata$L_RATIO)
k  <- rockchalk::kurtosis(mydata$L_RATIO, excess = FALSE)
k2 <- rockchalk::kurtosis(mydata$L_RATIO)

par(mfrow = c(1, 3), oma = c(0, 0, 2, 0))
hist(mydata$L_RATIO, col = "dodgerblue", xlab = "Log Ratio",
     main = "Histogram of Log Ratio")
qqnorm(mydata$L_RATIO, main = "Q-Q Plot",
       col = ifelse(mydata$L_RATIO %in% c(boxplot.stats(mydata$L_RATIO)$out), 
                    "darkorange", "dodgerblue"))
qqline(mydata$L_RATIO)
boxplot(mydata$L_RATIO ~ mydata$CLASS,
        main = "Boxplots of Log Ratio by Class",
        ylab = "Frequency",
        xlab = "Log Ratio Distribution by Class", 
        beside = TRUE,
        col = 'dodgerblue',
        pars = list(outcol = "darkorange"))
txt <- c("Skewness:", round(s, 2), 
         "Kurtosis:", round(k, 2), 
         "Excess kurtosis:", round(k2, 2))
mtext(paste(txt, sep=" ", collapse="  "), outer = TRUE, cex = 1)

```

(1)(c) Test the homogeneity of variance across classes using *bartlett.test()* (Kabacoff Section 9.2.2, p. 222). 

```{r Part_1c}
x <- bartlett.test(RATIO ~ CLASS, data = mydata)
y <- bartlett.test(L_RATIO ~ CLASS, data = mydata)
print(list('Ratio:', x, 'Log Ratio:', y))

```

**Essay Question: Based on steps 1.a, 1.b and 1.c, which variable RATIO or L_RATIO exhibits better conformance to a normal distribution with homogeneous variances across age classes?  Why?** 

***Answer: L_RATIO demonstrates improved adherence to a normal distribution, characterized by consistent variances across age classes. This is evidenced by reduced skewness in the histogram, a more uniform distribution of outliers in both QQ plots and boxplots, diminished skewness in the QQ plot, and the inability to reject the null hypothesis according to the Bartlett test for homogeneity of variances. ***


***#### Section 2 (10 points) ####***

(2)(a) Perform an analysis of variance with *aov()* on L_RATIO using CLASS and SEX as the independent variables (Kabacoff chapter 9, p. 212-229). Assume equal variances. Perform two analyses. First, fit a model with the interaction term CLASS:SEX. Then, fit a model without CLASS:SEX. Use *summary()* to obtain the analysis of variance tables (Kabacoff chapter 9, p. 227).

```{r Part_2a}
summary(aov(L_RATIO ~ CLASS + SEX + CLASS:SEX, data = mydata))
summary(aov(L_RATIO ~ CLASS + SEX, data = mydata))

```

**Essay Question:  Compare the two analyses.  What does the non-significant interaction term suggest about the relationship between L_RATIO and the factors CLASS and SEX?**

***Answer: The introduction of the interaction term had minimal impact and lacked statistical significance. Although the primary effects of CLASS and SEX are statistically significant in a model involving L_RATIO, the interaction between these two variables does not achieve statistical significance. ***

(2)(b) For the model without CLASS:SEX (i.e. an interaction term), obtain multiple comparisons with the *TukeyHSD()* function. Interpret the results at the 95% confidence level (*TukeyHSD()* will adjust for unequal sample sizes). 

```{r Part_2b}
TukeyHSD(aov(L_RATIO ~ CLASS + SEX, data = mydata))

```

**Additional Essay Question:  first, interpret the trend in coefficients across age classes. What is this indicating about L_RATIO?  Second, do these results suggest male and female abalones can be combined into a single category labeled as 'adults?' If not, why not?**

***Answer: The sole coefficient that appears to lack significance within the age classes pertains to the relationship between A2 and A1. Apart from this exception, all other coefficients demonstrate statistical significance. The p-value associated with the categories Infant and Male or Female suggests the rejection of the null hypothesis, indicating that the groups Male and Female can be effectively merged into a unified category termed "Adults" ***


***####  Section 3: (10 points) ####***

(3)(a1) Here, we will combine "M" and "F" into a new level, "ADULT". The code for doing this is given to you. For (3)(a1), all you need to do is execute the code as given.

```{r Part_3a1}
# Here, we show how to define the new variable TYPE using only base R functions:

mydata$TYPE <- factor(ifelse(mydata$SEX == "I", "I", "ADULT"))
table(mydata$TYPE)


```

(3)(a2)  Present side-by-side histograms of VOLUME. One should display infant volumes and, the other, adult volumes. 

```{r Part_3a2, fig.width = 12}

ggplot(mydata, aes(VOLUME, fill=TYPE)) +
  facet_grid(. ~ TYPE, space="free") +
  geom_histogram(color="black", size=.2, bins=30) + 
  scale_fill_manual(values=my_palette) +
  ggtitle("Histograms of Volume | Type") +
  theme(legend.position="none") +
  ylab("Frequency") +
  xlab("Volume")

```


**Essay Question: Compare the histograms.  How do the distributions differ? Are there going to be any difficulties separating infants from adults based on VOLUME?**

***Answer: The distribution of volumes for infants is right-skewed, concentrated toward lower values with potential outliers at higher volumes. In contrast, the distribution for adults appears more normal, with a predominant concentration above 300. The majority of infants' volumes are below 300. This observation implies that volume plays a significant role in distinguishing abalones, indicating its relevance in the separation process. ***

(3)(b) Create a scatterplot of SHUCK versus VOLUME and a scatterplot of their base ten logarithms, labeling the variables as L_SHUCK and L_VOLUME. Please be aware the variables, L_SHUCK and L_VOLUME, present the data as orders of magnitude (i.e. VOLUME = 100 = 10^2 becomes L_VOLUME = 2). Use color to differentiate CLASS in the plots. Repeat using color to differentiate by TYPE. 

```{r Part_3b, fig.width = 12, fig.height = 8}

mydata$L_SHUCK <- log10(mydata$SHUCK)
mydata$L_VOLUME <- log10(mydata$VOLUME)

s_v <- ggplot(data = mydata, aes(x = VOLUME, y = SHUCK)) + 
  geom_point(aes(color = CLASS),size = 1) + 
  ggtitle(bquote('Shuck Weight (grams) | Volume ('~cm^3~')')) +
  scale_color_manual(values=my_palette) +
  ylab("Shuck Weight (grams)") +
  xlab(bquote('Volume ('~cm^3~')'))

ls_lv <- ggplot(data = mydata, aes(x = L_VOLUME, y = L_SHUCK)) + 
  geom_point(aes(color = CLASS),size = 1) + 
  ggtitle('L_Shuck Weight | L_Volume ') +
  scale_color_manual(values=my_palette) +
  ylab("L_Shuck Weight") +
  xlab('L_Volume')

s_v_t <- ggplot(data = mydata, aes(x = VOLUME, y = SHUCK)) + 
  geom_point(aes(color = TYPE),size = 1) + 
  ggtitle(bquote('Shuck Weight (grams) | Volume ('~cm^3~')')) +
  scale_color_manual(values=my_palette) +
  ylab("Shuck Weight (grams)") +
  xlab(bquote('Volume ('~cm^3~')'))

ls_lv_t <- ggplot(data = mydata, aes(x = L_VOLUME, y = L_SHUCK)) + 
  geom_point(aes(color = TYPE),size = 1) + 
  ggtitle('L_Shuck Weight (grams) | L_Volume') +
  scale_color_manual(values=my_palette) +
  ylab("L_Shuck Weight") +
  xlab('L_Volume')

grid.arrange(s_v, ls_lv, s_v_t, ls_lv_t, nrow=2)

```

**Additional Essay Question:  Compare the two scatterplots. What effect(s) does log-transformation appear to have on the variability present in the plot?  What are the implications for linear regression analysis? Where do the various CLASS levels appear in the plots? Where do the levels of TYPE appear in the plots?**

***Answer: The initial distinction evident in the scatterplot lies in the reduced variability observed in the logarithmic rendition. This tighter fit to the regression line lends a more coherent pattern, particularly in illustrating the logical trend of weight augmentation with shuck increase. Consequently, this enhancement contributes to fortifying the linear regression for both Class and Sex variables. Examining the classes on the scatterplot, A1-2 occupies the bottom left, A3 is positioned at the graph's center, and A4+A5 resides in the top right. This spatial arrangement underscores the association that older abalones tend to exhibit greater weight. ***


***####   Section 4: (5 points) ####***

(4)(a1) Since abalone growth slows after class A3, infants in classes A4 and A5 are considered mature and candidates for harvest. You are given code in (4)(a1) to reclassify the infants in classes A4 and A5 as ADULTS. 

```{r Part_4a1}

mydata$TYPE[mydata$CLASS == "A4" | mydata$CLASS == "A5"] <- "ADULT"
table(mydata$TYPE)

```

(4)(a2) Regress L_SHUCK as the dependent variable on L_VOLUME, CLASS and TYPE (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2 and Black Section 14.2). Use the multiple regression model: L_SHUCK ~ L_VOLUME + CLASS + TYPE. Apply *summary()* to the model object to produce results.

```{r Part_4a2}

model <- lm(L_SHUCK ~ L_VOLUME + CLASS + TYPE, data=mydata)
summary(model)

```

**Essay Question:  Interpret the trend in CLASS levelcoefficient estimates? (Hint:  this question is not asking if the estimates are statistically significant. It is asking for an interpretation of the pattern in these coefficients, and how this pattern relates to the earlier displays).**

***Answer: The calculated coefficients indicate a more pronounced decline in L_SHUCK as the CLASS value increases. When considered alongside the earlier charts, this implies that L_SHUCK experiences a more substantial increase in the lower CLASS categories, after which it tends to plateau or even decrease on average as CLASS values rise. ***

**Additional Essay Question:  Is TYPE an important predictor in this regression? (Hint:  This question is not asking if TYPE is statistically significant, but rather how it compares to the other independent variables in terms of its contribution to predictions of L_SHUCK for harvesting decisions.)  Explain your conclusion.**

***Answer: TYPE demonstrates less significance compared to the majority of CLASS categories and is notably less important than L_VOLUME. This observation suggests that TYPE might not contribute substantially to predicting L_SHUCK, and conversely, the predictive power of L_SHUCK may not strongly influence the determination of TYPE.  ***

-----

The next two analysis steps involve an analysis of the residuals resulting from the regression model in (4)(a) (Kabacoff Section 8.2.4, p. 178-186, the Data Analysis Video #2).

-----

***#### Section 5: (5 points) ####***

(5)(a) If "model" is the regression object, use model$residuals and construct a histogram and QQ plot. Compute the skewness and kurtosis. Be aware that with 'rockchalk,' the kurtosis value has 3.0 subtracted from it which differs from the 'moments' package. 

```{r Part_5am, fig.width = 12}

s  <- rockchalk::skewness(model$residuals)
k  <- rockchalk::kurtosis(model$residuals, excess = FALSE)
k2 <- rockchalk::kurtosis(model$residuals)

par(mfrow = c(1, 2), oma = c(0, 0, 2, 0))
hist(model$residuals, col = "royalblue", xlab = "Residuals",
     main = "Histogram of Residuals")
qqnorm(model$residuals, main = "Q-Q Plot",
       col = ifelse(model$residuals %in% c(boxplot.stats(model$residuals)$out), 
                    "darkorchid", "royalblue"))
qqline(model$residuals)
txt <- c("Skewness:", round(s, 2), 
         "Kurtosis:", round(k, 2), 
         "Excess kurtosis:", round(k2, 2))
mtext(paste(txt, sep=" ", collapse="  "), outer = TRUE, cex = 1)

```

(5)(b) Plot the residuals versus L_VOLUME, coloring the data points by CLASS and, a second time, coloring the data points by TYPE. Keep in mind the y-axis and x-axis may be disproportionate which will amplify the variability in the residuals. Present boxplots of the residuals differentiated by CLASS and TYPE (These four plots can be conveniently presented on one page using *par(mfrow..)* or *grid.arrange()*. Test the homogeneity of variance of the residuals across classes using *bartlett.test()* (Kabacoff Section 9.3.2, p. 222).  

```{r Part_5b, fig.width = 12, fig.height = 8}

mydata$RESIDUALS <- model$residuals

gridExtra::grid.arrange(
  ggplot(mydata, aes(L_VOLUME, RESIDUALS)) + theme_bw() + 
    geom_point(aes(fill = CLASS), shape = 23),
  ggplot(mydata, aes(L_VOLUME, RESIDUALS)) + theme_bw() + 
    geom_point(aes(fill = TYPE), shape = 23),
  ggplot(mydata, aes(CLASS, RESIDUALS)) + theme_bw() +
    geom_boxplot(aes(fill = CLASS)),
  ggplot(mydata, aes(TYPE, RESIDUALS)) + theme_bw() +
    geom_boxplot(aes(fill = TYPE))
)
bartlett.test(RESIDUALS ~ CLASS, data = mydata)

```

**Essay Question:  What is revealed by the displays and calculations in (5)(a) and (5)(b)? Does the model 'fit'?  Does this analysis indicate that L_VOLUME, and ultimately VOLUME, might be useful for harvesting decisions? Discuss.**  

***Answer: The residuals exhibit a relatively even distribution and proximity to zero on both sides. While there isn't a distinctly defined pattern or shape, the presence of a large cluster on the right and a more widely dispersed scatter on the left in relation to VOLUME suggests that there may be room for improvement in the model. Addressing these patterns could potentially enhance the model's predictive performance.  ***

-----

Harvest Strategy:

There is a tradeoff faced in managing abalone harvest. The infant population must be protected since it represents future harvests. On the other hand, the harvest should be designed to be efficient with a yield to justify the effort. This assignment will use VOLUME to form binary decision rules to guide harvesting. If VOLUME is below a "cutoff" (i.e. a specified volume), that individual will not be harvested. If above, it will be harvested. Different rules are possible.The Management needs to make a decision to implement 1 rule that meets the business goal.

The next steps in the assignment will require consideration of the proportions of infants and adults harvested at different cutoffs. For this, similar "for-loops" will be used to compute the harvest proportions. These loops must use the same values for the constants min.v and delta and use the same statement "for(k in 1:10000)."  Otherwise, the resulting infant and adult proportions cannot be directly compared and plotted as requested. Note the example code supplied below.

-----

***#### Section 6: (5 points) ####***

(6)(a) A series of volumes covering the range from minimum to maximum abalone volume will be used in a "for loop" to determine how the harvest proportions change as the "cutoff" changes. Code for doing this is provided.

```{r Part_6a}

idxi <- mydata$TYPE == "I"
idxa <- mydata$TYPE == "ADULT"

max.v <- max(mydata$VOLUME)
min.v <- min(mydata$VOLUME)
delta <- (max.v - min.v)/10000
prop.infants <- numeric(10000)
prop.adults <- numeric(10000)
volume.value <- numeric(10000)

total.infants <- sum(idxi)  
total.adults <- sum(idxa)

for (k in 1:10000) { 
	value <- min.v + k*delta
	volume.value[k] <- value
	prop.infants[k] <- sum(mydata$VOLUME[idxi] <= value)/total.infants
	prop.adults[k] <-  sum(mydata$VOLUME[idxa] <= value)/total.adults
}

```

(6)(b) Our first "rule" will be protection of all infants. We want to find a volume cutoff that protects all infants, but gives us the largest possible harvest of adults. We can achieve this by using the volume of the largest infant as our cutoff. You are given code below to identify the largest infant VOLUME and to return the proportion of adults harvested by using this cutoff. You will need to modify this latter code to return the proportion of infants harvested using this cutoff. Remember that we will harvest any individual with VOLUME greater than our cutoff.

```{r Part_6b}
# Largest infant volume
c1_l = (max_inf_vol <- max(mydata$VOLUME[mydata$TYPE == "I"]))  # [1] 526.6383
cat("Largest infant volume: ", c1_l)

# Proportion of adults harvested
c1_a <- sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > max_inf_vol) / total.adults
cat("Proportion of adults harvested: ", c1_a)  # [1] 0.2476573

# Add code to calculate the proportion of infants harvested
c1_i <- sum(mydata$VOLUME[mydata$TYPE == "I"] > max_inf_vol) /
  total.infants
cat("Proportion of infants harvested: ", c1_i)

# If we use the largest infant volume, we harvest approximately 24.8% of adults and 0%,
# as expected, of infants.

```

(6)(c) Our next approaches will look at what happens when we use the median infant and adult harvest VOLUMEs. Using the median VOLUMEs as our cutoffs will give us (roughly) 50% harvests. We need to identify the median volumes and calculate the resulting infant and adult harvest proportions for both.

```{r Part_6c}
# Add code to determine the median infant volume:
med_inf_vol <- median(mydata$VOLUME[mydata$TYPE == "I"])

# Add code to calculate the proportion of infants harvested
c2_i <- sum(mydata$VOLUME[mydata$TYPE == "I"] > med_inf_vol) /
  total.infants
cat("Proportion of infants harvested: ", c2_i)

# Add code to calculate the proportion of adults harvested
c2_a <- sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_inf_vol) /
  total.adults
cat("Proportion of adults harvested: ", c2_a)

# If we use the median infant volume as our cutoff, we harvest almost 50% of our infants
# and a little more than 93% of our adults.


# Add code to determine the median adult volume:
med_ad_vol <- median(mydata$VOLUME[mydata$TYPE == "ADULT"])

# Add code to calculate the proportion of infants harvested
c3_i <- sum(mydata$VOLUME[mydata$TYPE == "I"] > med_ad_vol) /
  total.infants
cat("Proportion of infants harvested: ", c3_i)

# Add code to calculate the proportion of adults harvested
c3_a <- sum(mydata$VOLUME[mydata$TYPE == "ADULT"] > med_ad_vol) /
  total.adults
cat("Proportion of adults harvested: ", c3_a)

# If we use the median adult volume as our cutoff, we harvest almost 50% of adults
# and approximately 2.4% of infants.

```

(6)(d) Next, we will create a plot showing the infant conserved proportions (i.e. "not harvested," the prop.infants vector) and the adult conserved proportions (i.e. prop.adults) as functions of volume.value. We will add vertical A-B lines and text annotations for the three (3) "rules" considered, thus far:  "protect all infants," "median infant" and "median adult." Your plot will have two (2) curves - one (1) representing infant and one (1) representing adult proportions as functions of volume.value - and three (3) A-B lines representing the cutoffs determined in (6)(b) and (6)(c).

```{r Part_6d, fig.width = 12, fig.height = 6}

ggplot() + 
  geom_line(aes(volume.value, prop.infants, color="forestgreen")) +
  geom_line(aes(volume.value, prop.adults, color="gold")) +
  geom_vline(xintercept = max_inf_vol, linetype=2) +
  geom_vline(xintercept = med_inf_vol, linetype=2) +
  geom_vline(xintercept = med_ad_vol, linetype=2) +
  annotate("text", label=paste("Median Infant = ",round(med_inf_vol, 2)), x=med_inf_vol+10, y=.25, angle=90) +
  annotate("text", label=paste("Median Adult = ",round(med_ad_vol, 2)), x=med_ad_vol+10, y=.25, angle=90) +
  annotate("text", label=paste("Largest Infant = ",round(max_inf_vol, 2)), x=max_inf_vol+10, y=.25, angle=90) +
  ggtitle("Proportion of Adults and Infants Protected") +
  xlab('Volume ('~cm^3~')') +
  ylab("Proportion") +
  scale_color_manual(values = c("forestgreen", "gold"), name="Legend", labels=c("Proportion of Infants", "Proportion of Adults"))


```

**Essay Question:  The two 50% "median" values serve a descriptive purpose illustrating the difference between the populations. What do these values suggest regarding possible cutoffs for harvesting?** 

***Answer: The median volume for adults (384.56 cm³) is predictably higher than that for infants (133.82 cm³), indicating that using the median adult as the cutoff point would provide protection for more abalones, encompassing both infants and adults. If the objective is to conserve a greater number of abalones, adopting the more conservative median adult value as the cutoff is advisable. Conversely, if a more aggressive harvesting strategy is sought, opting for the median infant value as the cutoff could be considered. ***

-----


More harvest strategies:

This part will address the determination of a cutoff volume.value corresponding to the observed maximum difference in harvest percentages of adults and infants. In other words, we want to find the volume value such that the vertical distance between the infant curve and the adult curve is maximum. To calculate this result, the vectors of proportions from item (6) must be used. These proportions must be converted from "not harvested" to "harvested" proportions by using (1 - prop.infants) for infants, and (1 - prop.adults) for adults. The reason the proportion for infants drops sooner than adults is that infants are maturing and becoming adults with larger volumes.

-----

***#### Section 7: (10 points)  ####***

(7)(a) Evaluate a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value. Compare to the 50% "split" points determined in (6)(a). There is considerable variability present in the peak area of this plot. The observed "peak" difference may not be the best representation of the data. One solution is to smooth the data to determine a more representative estimate of the maximum difference.

```{r Part_7a}

difference <- (1 - prop.adults) - (1 - prop.infants)
ggplot() + 
  geom_line(aes(volume.value, difference), color=brewer.pal(name="GnBu", 9)[6]) +
  ggtitle("Difference in Proportions Harvested") +
  geom_vline(xintercept = med_inf_vol, linetype=2) +
  geom_vline(xintercept = med_ad_vol, linetype=2) +
  annotate("text", label=paste("Median Infant = ",round(med_inf_vol, 2)), x=med_inf_vol+20, y=.2, angle=90) +
  annotate("text", label=paste("Median Adult = ",round(med_ad_vol, 2)), x=med_ad_vol+20, y=.2, angle=90) +
  xlab("Volume") +
  ylab("Difference in Proportions Harvested")

```

(7)(b) Since curve smoothing is not studied in this course, code is supplied below. Execute the following code to create a smoothed curve to append to the plot in (a). The procedure is to individually smooth (1-prop.adults) and (1-prop.infants) before determining an estimate of the maximum difference. 

```{r Part_7b}

y.loess.a <- loess(1 - prop.adults ~ volume.value, span = 0.25,
	family = c("symmetric"))
y.loess.i <- loess(1 - prop.infants ~ volume.value, span = 0.25,
	family = c("symmetric"))
smooth.difference <- predict(y.loess.a) - predict(y.loess.i)

```

(7)(c) Present a plot of the difference ((1 - prop.adults) - (1 - prop.infants)) versus volume.value with the variable smooth.difference superimposed. Determine the volume.value corresponding to the maximum smoothed difference (Hint:  use *which.max()*). Show the estimated peak location corresponding to the cutoff determined.

Include, side-by-side, the plot from (6)(d) but with a fourth vertical A-B line added. That line should intercept the x-axis at the "max difference" volume determined from the smoothed curve here.

```{r Part_7c, fig.width = 12, fig.height = 6}

max <- volume.value[which.max(smooth.difference)]

diff <- ggplot() + 
  geom_line(aes(volume.value, difference), color=brewer.pal(name="GnBu", 9)[6]) +
  geom_line(aes(volume.value, smooth.difference), color=brewer.pal(name="GnBu", 9)[8], linetype=2) +
  geom_vline(xintercept = max) +
  annotate("text", label=paste("Volume = ",round(max, 2)), x=max+40, y=.2, angle=90) +
  ggtitle("Difference in Proportions Harvested") +
  xlab("Volume") +
  ylab("Difference in Proportions Harvested")

prop <- ggplot() + 
  geom_line(aes(volume.value, prop.infants, color="black")) +
  geom_line(aes(volume.value, prop.adults, color="green")) +
  geom_vline(xintercept = max_inf_vol, linetype=2) +
  geom_vline(xintercept = med_inf_vol, linetype=2) +
  geom_vline(xintercept = med_ad_vol, linetype=2) +
  geom_vline(xintercept = max, linetype=2) +
  annotate("text", label=paste("Median Infant = ",round(med_inf_vol, 2)), x=med_inf_vol+30, y=.25, angle=90) +
  annotate("text", label=paste("Median Adult = ",round(med_ad_vol, 2)), x=med_ad_vol+30, y=.25, angle=90) +
  annotate("text", label=paste("Largest Infant = ",round(max_inf_vol, 2)), x=max_inf_vol+30, y=.25, angle=90) +
  annotate("text", label=paste("Maximum Difference = ",round(max, 2)), x=max+30, y=.25, angle=90) +
  ggtitle("Proportion of Adults and Infants Protected") +
  xlab("Volume") +
  ylab("Proportion") +
  scale_color_manual(values=my_palette, name="Legend", labels=c("Proportion of Infants", "Proportion of Adults"))

grid.arrange(diff, prop, nrow=1)

```

(7)(d) What separate harvest proportions for infants and adults would result if this cutoff is used? Show the separate harvest proportions. We will actually calculate these proportions in two ways:  first, by 'indexing' and returning the appropriate element of the (1 - prop.adults) and (1 - prop.infants) vectors, and second, by simply counting the number of adults and infants with VOLUME greater than the vlume threshold of interest.

Code for calculating the adult harvest proportion using both approaches is provided.

```{r Part_7d}

c4_a <- (1 - prop.adults)[which.max(smooth.difference)]  # [1] 0.7416332
# OR,
c4_aa <- sum(mydata[mydata$TYPE == "ADULT", "VOLUME"] >
      volume.value[which.max(smooth.difference)]) / total.adults # [1] 0.7416332

cat("Proportion of adults harvested: ", c4_a)

# code to calculate the proportion of infants harvested
c4_i <- (1 - prop.infants)[which.max(smooth.difference)]
cat("Proportion of infants harvested: ", c4_i)

# code to output the "maximum smoothed difference" volume cutoff:
c4_diff <- volume.value[which.max(smooth.difference)]
cat("maximum smoothed difference: ", c4_diff)

```

-----

There are alternative ways to determine cutoffs. Two such cutoffs are described below.

-----

***####  Section 8: (10 points)  ####***

(8)(a) Harvesting of infants in CLASS "A1" must be minimized. The smallest volume.value cutoff that produces a zero harvest of infants from CLASS "A1" may be used as a baseline for comparison with larger cutoffs. Any smaller cutoff would result in harvesting infants from CLASS "A1."  

Compute this cutoff, and the proportions of infants and adults with VOLUME exceeding this cutoff. Code for determining this cutoff is provided. Show these proportions. You may use either the 'indexing' or 'count' approach, or both.

```{r Part_8a}

zero_a1_inf <- volume.value[volume.value > max(mydata[mydata$CLASS == "A1" &
  mydata$TYPE == "I", "VOLUME"])][1] # [1] 206.786
cat("cutoff: ", zero_a1_inf)

# code to calculate the proportion of infants harvested
c5_i <- sum(mydata[mydata$TYPE == "I", "VOLUME"] > zero_a1_inf) / total.infants
cat("Proportion of infants harvested: ", c5_i)

# code to calculate the proportion of adults harvested
c5_a <- sum(mydata[mydata$TYPE == "ADULT", "VOLUME"] > zero_a1_inf) / total.adults
cat("Proportion of adults harvested: ", c5_a)

```

(8)(b) Next, append one (1) more vertical A-B line to our (6)(d) graph. This time, showing the "zero A1 infants" cutoff from (8)(a). This graph should now have five (5) A-B lines:  "protect all infants," "median infant," "median adult," "max difference" and "zero A1 infants."

```{r Part_8b, fig.width = 12}

ggplot() + 
  geom_line(aes(volume.value, prop.infants, color="black")) +
  geom_line(aes(volume.value, prop.adults, color="green")) +
  geom_vline(xintercept = max_inf_vol, linetype=2) +
  geom_vline(xintercept = med_inf_vol, linetype=2) +
  geom_vline(xintercept = med_ad_vol, linetype=2) +
  geom_vline(xintercept = max, linetype=2) +
  geom_vline(xintercept = zero_a1_inf, linetype=2) +
  annotate("text", label=paste("Median Infant = ",round(med_inf_vol, 2)), x=med_inf_vol+10, y=.25, angle=90) +
  annotate("text", label=paste("Median Adult = ",round(med_ad_vol, 2)), x=med_ad_vol+10, y=.25, angle=90) +
  annotate("text", label=paste("Largest Infant = ",round(max_inf_vol, 2)), x=max_inf_vol+10, y=.25, angle=90) + 
  annotate("text", label=paste("Maximum Difference = ",round(max, 2)), x=max+10, y=.35, angle=90) +
  annotate("text", label=paste("Zero A1 Inf = ",round(zero_a1_inf, 2)), x=zero_a1_inf+10, y=.25, angle=90) +
  ggtitle("Proportion of Adults and Infants Protected") +
  xlab("Volume") +
  ylab("Proportion") +
  scale_color_manual(values=my_palette, name="Legend", labels=c("Proportion of Infants", "Proportion of Adults"))

```


***#### Section 9: (5 points) ####***

(9)(a) Construct an ROC curve by plotting (1 - prop.adults) versus (1 - prop.infants). Each point which appears corresponds to a particular volume.value. Show the location of the cutoffs determined in (6), (7) and (8) on this plot and label each. 

```{r Part_9, fig.width = 8.5}

ggplot() + 
  geom_line(aes(1-prop.infants, 1-prop.adults), color=brewer.pal(name="GnBu", 9)[6]) +
  geom_abline(slope=1, intercept=0, linetype=2) +
  ggtitle("ROC Curve of Adults vs Infants Harvested") +
  xlab("1-prop.infants") +
  ylab("1-prop.adults") +
  geom_point(aes(x=c1_i, y=c1_a), size=3, shape=1) +
  geom_point(aes(x=c2_i, y=c2_a), size=3, shape=1) +
  geom_point(aes(x=c3_i, y=c3_a), size=3, shape=1) +
  geom_point(aes(x=c4_i, y=c4_a), size=3, shape=1) +
  geom_point(aes(x=c5_i, y=c5_a), size=3, shape=1) +
  annotate("text", label=paste("Largest Infant = ",round(max_inf_vol, 2)), x=c1_i+.17, y=c1_a) +
  annotate("text", label=paste("Median Infant = ",round(med_inf_vol, 2)), x=c2_i+.2, y=c2_a) +
  annotate("text", label=paste("Median Adult = ",round(med_ad_vol, 2)), x=c3_i+.17, y=c3_a) +
  annotate("text", label=paste("Maximum Difference = ",round(max, 2)), x=c4_i+.25, y=c4_a) +
  annotate("text", label=paste("Zero A1 Inf = ",round(zero_a1_inf, 2)), x=c5_i+.2, y=c5_a)

```

(9)(b) Numerically integrate the area under the ROC curve and report your result. This is most easily done with the *auc()* function from the "flux" package.   Areas-under-curve, or AUCs, greater than 0.8 are taken to indicate good discrimination potential. 

```{r Part_9b}
paste("Area under ROC curve:", flux::auc(x = (1 - prop.infants), y = (1 - prop.adults)))

```


***#### Section 10: (10 points) ####***

(10)(a) Prepare a table showing each cutoff along with the following:
 	1) true positive rate (1-prop.adults,
 	2) false positive rate (1-prop.infants),
 	3) harvest proportion of the total population

To calculate the total harvest proportions, you can use the 'count' approach, but ignoring TYPE; simply count the number of individuals (i.e. rows) with VOLUME greater than a given threshold and divide by the total number of individuals in our dataset.
 	
```{r Part_10} 	

c1_t <- (c1_a*total.adults + c1_i*total.infants)/(total.adults+total.infants)
c2_t <- (c2_a*total.adults + c2_i*total.infants)/(total.adults+total.infants)
c3_t <- (c3_a*total.adults + c3_i*total.infants)/(total.adults+total.infants)
c4_t <- (c4_a*total.adults + c4_i*total.infants)/(total.adults+total.infants)
c5_t <- (c5_a*total.adults + c5_i*total.infants)/(total.adults+total.infants)

c1 <- c(max_inf_vol, c1_a, c1_i, c1_t)
c2 <- c(med_inf_vol, c2_a, c2_i, c2_t)
c3 <- c(med_ad_vol, c3_a, c3_i, c3_t)
c4 <- c(max, c4_a, c4_i, c4_t)
c5 <- c(zero_a1_inf, c5_a, c5_i, c5_t)
c <- rbind(c1, c2, c3, c4, c5)
colnames(c) <- c("Volume", "True Positive", "False Positive", "totalHarvest")
rownames(c) <- c("Protect All Infants", "Median Infants", "Median Adults", "Max Difference", "Zero A1 Infants")
kable(c)

```
 	
**Essay Question: Based on the ROC curve, it is evident a wide range of possible "cutoffs" exist. Compare and discuss the five cutoffs determined in this assignment.**   

***Answer: The five designated cut-off values offer a spectrum of potential harvesting strategies, ranging from aggressive to conservative. The median infant value signifies the most aggressive approach, resulting in the lowest overall preservation rate. Successive cut-off values, namely zero A1 infants, maximum infant-adult difference, median adult value, and largest infant, progressively preserve more abalone overall. However, the decision is not solely about preserving the least or the most abalone; the cut-off point also introduces a range of options regarding the proportion of adult to infant abalone preservation. Notably, the "maximum difference" cut-off yields the largest adult-to-infant harvesting proportion, prioritizing the preservation of adult abalones while restricting the number of infants harvested proportionally. ***



**Final Essay Question:  Assume you are expected to make a presentation of your analysis to the investigators How would you do so?  Consider the following in your answer:**

1. Would you make a specific recommendation or outline various choices and tradeoffs?
2. What qualifications or limitations would you present regarding your analysis?
3. If it is necessary to proceed based on the current analysis, what suggestions would you have for  implementation of a cutoff?  
4. What suggestions would you have for planning future abalone studies of this type? 

***Answer: 1. I would not recommend anything but i would highlight the choices presented in the table and emphasize on the false positive rate of the options. 2. I would present the distribution of the data, presence of outliers, the difficulty in correctly assessing the classes of abalones and highlight the limited scope of the analysis due to constraints presented by the quality of data collection methods. 3. If the analysis was to continue forward, i would recommend going with the most conservative approach. 4. I would suggest ways to collect better and more data - environmental, location, diet, etc. By collecting more and better data, the FPR could possible be reduced. ***