---
title: 'Data Analysis Assignment #1 (50 points total)'
author: "Sharma, Sachin"
output:
  html_document: default
---

```{r setup, include = FALSE}
# DO NOT ADD OR REVISE CODE HERE
knitr::opts_chunk$set(echo = FALSE, eval = TRUE)

```


Submit both the .Rmd and .html files for grading. You may remove the instructions and example problem above, but do not remove the YAML metadata block or the first, "setup" code chunk.  Address the steps that appear below and answer all the questions. Be sure to address each question with code and comments as needed.  You may use either base R functions or ggplot2 for the visualizations.

-----

The following code chunk will:

(a) load the "ggplot2", "gridExtra" and "knitr" packages, assuming each has been installed on your machine,
(b) read-in the abalones dataset, defining a new data frame, "mydata,"
(c) return the structure of that data frame, and
(d) calculate new variables, VOLUME and RATIO.

Do not include package installation code in this document. Packages should be installed via the Console or 'Packages' tab. You will also need to download the abalones.csv from the course site to a known location on your machine. Unless a *file.path()* is specified, R will look to directory where this .Rmd is stored when knitting.

```{r analysis_setup1, message = FALSE, warning = FALSE}

# a) Load the ggplot2 and gridExtra packages.

library(ggplot2)
library(gridExtra)
library(knitr)
library(RColorBrewer)

# b) Use read.csv() to read the abalones.csv into R, assigning the data frame to "mydata."

mydata <- read.csv("abalones.csv", sep = ",", stringsAsFactors = TRUE)


# c) Use the str() function to verify the structure of "mydata." You should have 1036 observations
# of eight variables.

str(mydata)

# d) Define two new variables, VOLUME and RATIO. Use the following statements to define VOLUME and
# RATIO as variables appended to the data frame "mydata."

mydata$VOLUME <- mydata$LENGTH * mydata$DIAM * mydata$HEIGHT
mydata$RATIO <- mydata$SHUCK / mydata$VOLUME

```

-----

### Test Items starts from here - There are 6 sections - Total 50 points

***##### Section 1: (6 points) Summarizing the data.***

(1)(a) (1 point) Use *summary()* to obtain and present descriptive statistics from mydata.  Use *table()* to present a frequency table using CLASS and RINGS. There should be 115 cells in the table you present.  

```{r Part_1a}
summary(mydata)
table(mydata$CLASS, mydata$RINGS)
```

**Question (1 point):  Briefly discuss the variable types and distributional implications such as potential skewness and outliers.**

***Answer: The distribution of Female, Male, and Infants seems to be equal. However, there could be some outliers present in the data due to larger abalones, as indicated by the maximum values of Whole Wt, Volume, and Shuck variables.***

(1)(b) (1 point) Generate a table of counts using SEX and CLASS. Add margins to this table (Hint: There should be 15 cells in this table plus the marginal totals. Apply *table()* first, then pass the table object to *addmargins()* (Kabacoff Section 7.2 pages 144-147)).  Lastly, present a barplot of these data; ignoring the marginal totals.  

```{r Part_1b}
t <- table(mydata$SEX, mydata$CLASS)
addmargins(t)

freq <- data.frame(t)
colnames(freq) <- c('Sex', 'Class', 'Frequency')
ggplot(freq, aes(x=Class, y=Frequency, fill=Sex, group=Sex)) + 
  geom_bar(stat="identity", position=position_dodge()) +
  ggtitle("Abalone by Sex - Class")
```

**Essay Question (2 points):  Discuss the sex distribution of abalones.  What stands out about the distribution of abalones by CLASS?**

***Answer: It is not surprising that the youngest Abalones (A1) belong to the Infant category "I". The number of abalones increases until mid-life (A3), after which it steadily declines. It is noteworthy that the proportion of male and female abalones becomes almost equal as they age (A5 being the oldest).***


(1)(c) (1 point) Select a simple random sample of 200 observations from "mydata" and identify this sample as "work." Use *set.seed(123)* prior to drawing this sample. Do not change the number 123. Note that *sample()* "takes a sample of the specified size from the elements of x." We cannot sample directly from "mydata." Instead, we need to sample from the integers, 1 to 1036, representing the rows of "mydata." Then, select those rows from the data frame (Kabacoff Section 4.10.5 page 87). 

Using "work", construct a scatterplot matrix of variables 2-6 with *plot(work[, 2:6])* (these are the continuous variables excluding VOLUME and RATIO). The sample "work" will not be used in the remainder of the assignment. 

```{r Part_1c}
set.seed(123)
srs <- sample(seq(1:1036), 200)
work <- mydata[srs,]
plot(work[, 2:6], cex=.6, lwd=.4)

```

-----

***##### Section 2: (5 points) Summarizing the data using graphics.***

(2)(a) (1 point) Use "mydata" to plot WHOLE versus VOLUME. Color code data points by CLASS.

```{r Part_2a}

ggplot(data = mydata, aes(x = VOLUME, y = WHOLE)) + 
  geom_point(aes(color = CLASS),size = 1) + 
  ggtitle(bquote('Whole weight (grams) vs Volume ('~cm^3~')')) +
  scale_color_brewer(palette="GnBu") +
  ylab("Whole weight (grams)") +
  xlab(bquote('Volume ('~cm^3~')'))
```


(2)(b) (2 points) Use "mydata" to plot SHUCK versus WHOLE with WHOLE on the horizontal axis. Color code data points by CLASS. As an aid to interpretation, determine the maximum value of the ratio of SHUCK to WHOLE.  Add to the chart a straight line with zero intercept using this maximum value as the slope of the line. If you are using the 'base R' *plot()* function, you may use *abline()* to add this line to the plot. Use *help(abline)* in R to determine the coding for the slope and intercept arguments in the functions. If you are using ggplot2 for visualizations, *geom_abline()* should be used.

```{r Part_2b}
ggplot(data = mydata, aes(x = WHOLE, y = SHUCK)) + 
  geom_point(aes(color = CLASS),size = 1,) +
  scale_color_brewer(palette="GnBu") +
  ggtitle('Shuck weight (grams) vs Whole weight (grams)') +
  geom_abline(intercept=0, slope=max(mydata$SHUCK/mydata$WHOLE), size=.25, linetype=2) +
  ylab("Shuck weight (grams)") +
  xlab('Whole weight (grams)') 

```

**Essay Question (2 points):  How does the variability in this plot differ from the plot in (a)?  Compare the two displays.  Keep in mind that SHUCK is a part of WHOLE.  Consider the location of the different age classes.**

***Answer: Plot (a) displays more variability than plot (b). The presence of two maximum outliers linked to A1/Infants in plot (a) could be distorting the outcomes. Furthermore, plot (b) represents the maximum ratio of shuck to whole by a straight line, but most of the abalones lie below this line, indicating that waiting for older abalones (A3 - A5) may not necessarily result in greater shuck weight.***

-----

***### Section 3: (8 points) Getting insights about the data using graphs.***

(3)(a) (2 points) Use "mydata" to create a multi-figured plot with histograms, boxplots and Q-Q plots of RATIO differentiated by sex. This can be done using *par(mfrow = c(3,3))* and base R or *grid.arrange()* and ggplot2. The first row would show the histograms, the second row the boxplots and the third row the Q-Q plots. Be sure these displays are legible.  

```{r Part_3a}
a <- ggplot(mydata, aes(RATIO, fill=SEX)) +
  facet_grid(. ~ SEX, space="free") +
  geom_histogram(bins=14, color="black", size=.2) +
  ggtitle("Histograms of Ratio") +
  theme(legend.position="none", axis.title.x=element_blank()) +
  ylab("Frequency")
  
b <- ggplot(mydata, aes(y=RATIO, fill=SEX)) +
  facet_grid(cols = vars(SEX)) +
  geom_boxplot(color="black", size=.2, outlier.size=2, outlier.shape=1) + 
  ggtitle("Boxplots of Ratio") +
  theme(legend.position="none", axis.title=element_blank())

c <- ggplot(mydata, aes(sample=RATIO, color=SEX)) +
  facet_grid(cols = vars(SEX)) +
  geom_qq(cex=2, shape=1) + 
  ggtitle("QQ Plots of Ratio") +
  geom_qq_line(color="black", cex=.5) +
  theme(legend.position="none") +
  ylab("Sample Quantities") +
  xlab("Theoretical Quantities")

grid.arrange(a, b, c, nrow=3)

```

**Essay Question (2 points): Compare the displays.  How do the distributions compare to normality? Take into account the criteria discussed in the sync sessions to evaluate non-normality.**

***Answer: The histogram indicates that the outcomes are not normally distributed and are skewed, possibly due to outliers within each sex. Moreover, the Q-Q plot for Infant and Female categories deviates from normal distribution, further confirming the non-normality of the distribution.***

(3)(b) (2 points) The boxplots in (3)(a) indicate that there are outlying RATIOs for each sex. *boxplot.stats()* can be used to identify outlying values of a vector. Present the abalones with these outlying RATIO values along with their associated variables in "mydata". Display the observations by passing a data frame to the kable() function. Basically, we want to output those rows of "mydata" with an outlying RATIO, but we want to determine outliers looking separately at infants, females and males.

```{r Part_3b}
i_out <- mydata[mydata$SEX == "I" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "I"])$out, ]
f_out <- mydata[mydata$SEX == "F" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "F"])$out, ]
m_out <- mydata[mydata$SEX == "M" & mydata$RATIO %in% boxplot.stats(x = mydata$RATIO[mydata$SEX == "M"])$out, ]
kable(rbind(i_out, f_out, m_out))

```

**Essay Question (2 points):  What are your observations regarding the results in (3)(b)?**

***Answer: There is an outlier in the AI/Infant class, indicating a higher number of rings (6), larger Whole Weight, and volume, which is more commonly associated with older abalones. Furthermore, the volume and whole weight decline as the abalones age.***

-----

***### Section 4: (8 points) Getting insights about possible predictors.***

(4)(a) (3 points) With "mydata," display side-by-side boxplots for VOLUME and WHOLE, each differentiated by CLASS There should be five boxes for VOLUME and five for WHOLE. Also, display side-by-side scatterplots:  VOLUME and WHOLE versus RINGS. Present these four figures in one graphic:  the boxplots in one row and the scatterplots in a second row. Base R or ggplot2 may be used.


```{r Part_4a}
a_v <- ggplot(mydata, aes(x=CLASS, y=VOLUME, fill=CLASS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1) + 
  scale_fill_brewer(palette="GnBu") +
  ggtitle("Boxplots of Volume | Class") +
  theme(legend.position="none")

b_w <- ggplot(mydata, aes(x=CLASS, y=WHOLE, fill=CLASS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1) + 
  scale_fill_brewer(palette="GnBu") +
  ggtitle("Boxplots of Whole weight | Class") +
  theme(legend.position="none")

c_v <- ggplot(data = mydata, aes(x = RINGS, y = VOLUME)) + 
  geom_point(size = .5,) +
  scale_color_brewer(palette="GnBu") +
  ggtitle("Volume | Rings") +
  theme(legend.position="none")

d_w <- ggplot(data = mydata, aes(x = RINGS, y = WHOLE)) + 
  geom_point(size = .5,) +
  scale_color_brewer(palette="GnBu") +
  ggtitle("Whole weight | Rings") +
  theme(legend.position="none")

grid.arrange(a_v, b_w, c_v, d_w, nrow=2)

```

**Essay Question (5 points) How well do you think these variables would perform as predictors of age?  Explain.**

***Answer: The boxplot shows that for the older group of abalones, the median of A4 and A5 is almost the same. Hence, Whole Weight cannot be a reliable indicator of age. Moreover, the scatterplot would not be an effective tool to determine age, as there is more variability in Whole Weight and volume with an increase in the number of rings.***

-----

***### Section 5: (12 points) Getting insights regarding different groups in the data.***

(5)(a) (2 points) Use *aggregate()* with "mydata" to compute the mean values of VOLUME, SHUCK and RATIO for each combination of SEX and CLASS. Then, using *matrix()*, create matrices of the mean values. Using the "dimnames" argument within *matrix()* or the *rownames()* and *colnames()* functions on the matrices, label the rows by SEX and columns by CLASS. Present the three matrices (Kabacoff Section 5.6.2, p. 110-111). The *kable()* function is useful for this purpose.  You do not need to be concerned with the number of digits presented.

```{r Part_5a}
agg <- aggregate(mydata[c('VOLUME', 'SHUCK', 'RATIO')], FUN=mean, by=list(mydata$SEX, mydata$CLASS))

a_v <- matrix(agg$VOLUME, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

b_s <- matrix(agg$SHUCK, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

c_r <- matrix(agg$RATIO, nrow=3, dimnames = list(list('F', 'I', 'M'), list('A1', 'A2', 'A3', 'A4', 'A5')))

kable(a_v, caption="Volume")
kable(b_s, caption="Shuck")
kable(c_r, caption="Ratio")

```

(5)(b) (3 points) Present three graphs. Each graph should include three lines, one for each sex. The first should show mean RATIO versus CLASS; the second, mean VOLUME versus CLASS; the third, mean SHUCK versus CLASS. This may be done with the 'base R' *interaction.plot()* function or with ggplot2 using *grid.arrange()*.

```{r Part_5b, fig.width = 9}
colnames(agg) <- c('Sex', 'Class', 'Volume', 'Shuck', 'Ratio')
a_r <- ggplot(data = agg, aes(x = Class, y = Ratio, group=Sex, color=Sex)) + 
  geom_line(size=1) +
  ggtitle("Mean Ratio | Class")

b_r <- ggplot(data = agg, aes(x = Class, y = Volume, group=Sex, color=Sex)) + 
  geom_line(size=1) +
  ggtitle("Mean Volume | Class")

c_r <- ggplot(data = agg, aes(x = Class, y = Shuck, group=Sex, color=Sex)) + 
  geom_line(size=1) +
  ggtitle("Mean Shuck weight | Class")

grid.arrange(a_r, b_r, c_r, nrow=3)

```

**Essay Question (2 points):  What questions do these plots raise?  Consider aging and sex differences.**

***Answer: In general, the mean shuck/volume ratio decreases with age (class), while both volume and shuck increase with age across all sexes. However, the plots raise several questions about the differences in shuck and volume between genders. For instance, why do females tend to be heavier and larger than males and infants? Additionally, why do females exhibit a slower growth rate than males and infants despite being larger? Is the small difference between female A1 and A2 (in terms of volume and shuck) indicative of slower development? Also, why do infants consistently weigh less and have smaller sizes than adults, but also have a lower shuck/volume ratio? Is this because infants genuinely have a lower shuck/volume ratio, or is it simply more challenging to determine the sex of an abalone with a lower shuck/volume ratio? Finally, why do both adults and infants tend to shrink (become lighter/smaller) between A4 and A5?***

5(c) (3 points) Present four boxplots using *par(mfrow = c(2, 2)* or *grid.arrange()*. The first line should show VOLUME by RINGS for the infants and, separately, for the adult; factor levels "M" and "F," combined. The second line should show WHOLE by RINGS for the infants and, separately, for the adults. Since the data are sparse beyond 15 rings, limit the displays to less than 16 rings. One way to accomplish this is to generate a new data set using subset() to select RINGS < 16.  Use ylim = c(0, 1100) for VOLUME and ylim = c(0, 400) for WHOLE.  If you wish to reorder the displays for presentation purposes or use ggplot2 go ahead.

```{r Part_5c}
a_v <- ggplot(subset(mydata, mydata$RINGS < 16 & mydata$SEX == "I"), aes(x=RINGS, y=VOLUME, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="steelblue") + 
  ggtitle("Infant Volume | Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 1000)) +
  ylab("Volume") +
  xlab("Rings")

b_v <- ggplot(subset(mydata, mydata$RINGS < 16 & (mydata$SEX == "M" | mydata$SEX == "F")), aes(x=RINGS, y=VOLUME, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="tomato") + 
  ggtitle("Adult Volume | Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 1000)) +
  ylab("Volume") +
  xlab("Rings")
  
c_w <- ggplot(subset(mydata, mydata$RINGS < 16 & mydata$SEX == "I"), aes(x=RINGS, y=WHOLE, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="darkolivegreen") + 
  ggtitle("Infant Whole weight | Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 400)) +
  ylab("Whole weight") +
  xlab("Rings")

d_w <- ggplot(subset(mydata, mydata$RINGS < 16 & (mydata$SEX == "M" | mydata$SEX == "F")), aes(x=RINGS, y=WHOLE, fill=RINGS, group=RINGS)) +
  geom_boxplot(color="black", size=.1, outlier.shape=1, fill="orchid") + 
  ggtitle("Adult Whole weight | Rings") +
  theme(legend.position = "none") + 
  coord_cartesian(ylim = c(0, 400)) +
  ylab("Whole weight") +
  xlab("Rings")

grid.arrange(a_v, b_v, c_w, d_w, nrow=2)

```

**Essay Question (2 points):  What do these displays suggest about abalone growth?  Also, compare the infant and adult displays.  What differences stand out?**

***Answer: The visual representations indicate that abalones attain most of their volume when they reach adulthood rather than during their infancy. Additionally, despite restricting the number of rings to less than 16, the graphs demonstrate that there is minimal disparity in both Whole Weight and Volume between adults and infants.***

-----

***### Section 6: (11 points) Conclusions from the Exploratory Data Analysis (EDA).***


**Conclusions**


**Essay Question 1) (5 points)	Based solely on these data, what are plausible statistical reasons that explain the failure of the original study? Consider to what extent physical measurements may be used for age prediction.**

***Answer: The presence of outliers and the failure to address them may explain the shortcomings of the original study. While using physical measurements as an indicator of age can be a useful starting point for further investigation and experimentation, drawing conclusions solely based on observational measurements is not recommended.***

**Essay Question 2) (3 points) Do not refer to the abalone data or study.  If you were presented with an overall histogram and summary statistics from a sample of some population or phenomenon and no other information, what questions might you ask before accepting them as representative of the sampled population or phenomenon?**

***Answer: I am interested in learning more about the sample and the population. Can you provide details on the sampling methods used? Was a simple random sample employed, or did the study use a non-random sampling method? How was the sampling frame constructed? Also, when was the sample collected, and is the data outdated? Additionally, what is the size of the sample and how does it compare to the population size? Are there any known population parameters or can they be estimated? Are there any biases or other known influential factors that might influence how representative the sample is of the population? There are numerous factors that could affect whether or not the sample is representative of the population being studied.***

**Essay Question 3)  (3 points)	Do not refer to the abalone data or study.  What do you see as difficulties analyzing data derived from observational studies? Can causality be determined?  What might be learned from such studies?**

***Answer: Analyzing data from observational studies can be challenging due to the involvement of numerous factors. Human error and bias, whether during the study's design, data collection, or result analysis, can significantly influence observational studies. It is effortless for an observational study to neglect or overlook various explanatory variables that may impact the response variable. Without a control group, an observational study can only establish correlation at best. It is impossible to establish causality from an observational study; however, it can identify variables of interest that require further investigation to determine causality.***
